---
title: '`voice` vignette - The R part'
subtitle: 'version `r packageVersion("voice")`'
author: 'Filipe J. Zabala'
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{`voice` vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ini0 <- Sys.time()
```

## 0. Installation
https://github.com/filipezabala/voice

## 1. Extract features
### Load packages
```{r, message=FALSE, warning=FALSE}
# packs
library(voice)
library(tidyverse)
library(music)
library(gm)
```

### Setting directories and removing old ones
```{r}
dirs <- c('vaDir', 'wavDir', 'mp3Dir', 'musicxmlDir')
vaDir <- '~/Downloads/voiceAudios'
for(i in 1:length(dirs)){
  if(i>1){ assign(dirs[i], paste0(get(dirs[1]), '/', dirs[i])) }
  ifelse(!dir.exists(get(dirs[i])), dir.create(get(dirs[i])), 'Directory exists!')
}
cmd <- paste0('rm ', wavDir, '/*.*; ', 'rm ', mp3Dir, '/*.*')
system(cmd)
```


<!-- #### >>> To do #10 [SOLVED @ VERSION 0.0.0.9044 2021-09-12] -->
<!-- Automatize and test `autoDir` argument @ splitw.R.  <<< -->

<!-- #### >>> To do #20 [SOLVED @ VERSION 0.0.0.9044 2021-09-12] -->
<!-- Parameters fromWav, fromRttm admits either file or directory @ splitw.R. <<< -->

### Get audio
```{r}
url0 <- 'https://github.com/filipezabala/voiceAudios/blob/main/mp3/doremi.mp3?raw=true'
download.file(url0, paste0(mp3Dir, '/doremi.mp3'), mode = 'wb')
embedr::embed_audio(url0)
```

<!-- #### >>> To do #30 [SOLVED in 2021-11-07] -->
<!-- Insert a button to play audio files. <<< -->

<!-- #### >>> To do #40 -->
<!-- At `url0 <- 'https://github.com/filipezabala/voiceAudios/raw/main/bebezinho_2.005.wav'`, find out why the downloaded file is called `wav_bebezinho_2.005.wav`. <<< -->

### Convert mp3 to wav mono
```{r}
cmd <- 'cd ~/Downloads/voiceAudios/mp3Dir;
for i in *.[Mm][Pp]3; do ffmpeg -i "$i" -ac 1 "../wavDir/${i%.*}.wav"; done'
system(cmd)
```

### Extract features
```{r}
ef <- voice::extract_features(wavDir, features = c('f0','gain'),
                              round.to = 6, windowShift = 5)
ef
```
<!-- #### >>> To do #50 -->
<!-- Automatize and test `stereo2mono`. <<< -->

<!-- #### >>> To do #60 -->
<!-- Test different windowShift values.  <<< -->


## 2. Smooth
<!-- #### >>> To do #70 [SOLVED @ VERSION 0.0.0.9042 2021-09-12] -->
<!-- Write the function `smooth_df(x,k)`. <<< -->

```{r, warning=FALSE, message=FALSE}
odd <- seq(1,11,2)
ef_sm <- vector('list', length(odd))
names(ef_sm) <- paste0('k=', odd)
for(i in 1:length(ef_sm)){
  ef_sm[[i]] <- voice::smooth_df(ef, id = 3, k = i)
}
sapply(ef_sm, dim)
```

### Plot F0 smoothing
```{r, fig.height=6, fig.width=8}
par(mfrow = c(2,3))
for(i in 1:length(ef_sm)){
  plot(ef_sm[[i]]$F0, main = paste0('k = ', odd[i]))
}
```


<!-- ### Pool -->
<!-- ```{r} -->
<!-- (ef01 <- voice::conv_df(ef, .01)) # 1% -->
<!-- ``` -->
<!-- #### >>> To do -->
<!-- 1. Eliminate this step, assigning notes and durations after smooth with moving average.   -->
<!-- 2. Apply voice::duration over the smoothed data.   -->

## 3. Assign notes
Assign notes in Scientific Pitch Notation to F0.
```{r}
for(i in 1:length(ef_sm)){
  ef_sm[[i]]$F0_spn <- voice::notes(ef_sm[[i]]$F0)
}
```

<!-- #### >>> To do #80 [WORK IN PROGRESS] -->
<!-- Write plot_note function to show both frequency and notes in spn, midi and black formats. <<< -->

<!-- #### >>> To do #90 -->
<!-- Write a function to, given a frequency in Hz, convert in any SPN. <<< -->

<!-- #### >>> To do #100 -->
<!-- Implement and study major/minor chords identifier (consider `tabr` functions). <<< -->

<!-- #### >>> To do #110 -->
<!-- Study sequences (scales and arpeggios). <<< -->


## 4. Duration
```{r}
dur_spn <- vector('list', length(ef_sm))
names(dur_spn) <- paste0('k=', odd)
for(i in 1:length(ef_sm)){
  dur_spn[[i]] <- voice::duration(ef_sm[[i]]$F0_spn) 
}
```

<!-- ## 5. Distance -->
<!-- Distance in semitones -->
<!-- ```{r} -->
<!-- nd <- vector('list', length(ef_sm)) -->
<!-- for(i in 1:length(ef_sm)){ -->
<!--   nd[[i]] <- music::noteDistance(as.character(dur_spn[[i]]$note)) -->
<!--   print(summary(nd[[i]])) -->
<!-- } -->
<!-- nd[[1]] -->
<!-- ``` -->

<!-- #### >>> To do #120 -->
<!-- Extract and test microtones. (Giuliano's idea.) <<< -->

<!-- ### Play -->
<!-- ```{r} -->
<!-- music::playNote(note = as.character(dur.spn$note), duration = dur.spn$dur_line) -->
<!-- ``` -->
<!-- #### >>> To do -->
<!-- 1. Check version to play different durations.   -->

<!-- To do:   -->
<!-- 1. Solve in Linux:   -->

<!-- ```{r} -->
<!-- # ALSA lib pcm_dmix.c:1089:(snd_pcm_dmix_open) unable to open slave -->
<!-- # ALSA lib pcm.c:2642:(snd_pcm_open_noupdate) Unknown PCM cards.pcm.rear -->
<!-- # ALSA lib pcm.c:2642:(snd_pcm_open_noupdate) Unknown PCM cards.pcm.center_lfe -->
<!-- # ALSA lib pcm.c:2642:(snd_pcm_open_noupdate) Unknown PCM cards.pcm.side -->
<!-- # ALSA lib pcm_route.c:869:(find_matching_chmap) Found no matching channel map -->
<!-- # ALSA lib pcm_oss.c:377:(_snd_pcm_oss_open) Unknown field port -->
<!-- # ALSA lib pcm_oss.c:377:(_snd_pcm_oss_open) Unknown field port -->
<!-- # ALSA lib pulse.c:242:(pulse_connect) PulseAudio: Unable to connect: Connection refused -->
<!-- #  -->
<!-- # ALSA lib pulse.c:242:(pulse_connect) PulseAudio: Unable to connect: Connection refused -->
<!-- #  -->
<!-- # ALSA lib pcm_usb_stream.c:486:(_snd_pcm_usb_stream_open) Invalid type for card -->
<!-- # ALSA lib pcm_usb_stream.c:486:(_snd_pcm_usb_stream_open) Invalid type for card -->
<!-- # ALSA lib pcm_dmix.c:1089:(snd_pcm_dmix_open) unable to open slave -->
<!-- # ALSA lib pcm_dmix.c:1089:(snd_pcm_dmix_open) unable to open slave -->
<!-- ``` -->

## 5. Music sheet
### Compress to 1%
```{r}
# compress
ef_01 <- voice::conv_df(ef, 0.01, id = 3)

# assign notes
ef_01$F0_spn <- voice::notes(ef_01$F0, measure = 'spn')

# duration
nd_01 <- voice::duration(ef_01$F0_spn)

# gm by Renfei Mao
m01 <- gm::Music()
m01 <- m01 +
  gm::Meter(4, 4) +
  gm::Line(pitches = as.list(as.character(nd_01$note)),
           durations = as.list(nd_01$dur_line)) +
  Tempo(170)
show(m01, to = c('score', 'audio'))
```


## 6. Test your voice

https://codepen.io/aeewhite/pen/BjzbOL

<script>
<h2>Not Recording</h2>

<button class="notRec">Not Recording</button>

<br />
<br />
<h2>Recording</h2>

<button class="Rec">Recording</button>

<h2>Demo</h2>
<p>Click the button to start/stop recording</p>
<button id="recButton"></button>

</script>

```{r}
tm <- 2 # desired time in seconds
rate <- 8000 # sample recording rate in Hz
x <- rep(NA_real_, rate*tm)
a <- audio::record(x, rate, 1)
wait(a)
class(a)
play(x, rate = rate*2)
save.wave(a$data, 'test.WAV')
# aa <- load.wave('test.WAV')
# play(aa, rate = rate*2)
a2 <- tuneR::readWave('test.WAV')
embedr::embed_audio(a2)

```


<!-- #### >>> To do #130 -->
<!-- Automate `pitches` and `durations` at `gm::Line` to capture spn/midi from `duration` data frame. <<< -->


<!-- ### Export musicxml file -->
<!-- ```{r} -->
<!-- gm::export(m01, musicxmlDir, 'bebezinho_01', 'musicxml') -->
<!-- gm::export(m21, musicxmlDir, 'bebezinho_21', 'musicxml') -->
<!-- ``` -->


<!-- ```{r} -->
<!-- Sys.time()-ini0 -->
<!-- ``` -->

