---
title: '`voice` vignette'
subtitle: 'version `r packageVersion("voice")`'
author: 'Filipe J. Zabala'
date: "`r format(Sys.time(), '%Y-%m-%d')`"
# date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{`voice` vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
ini0 <- Sys.time()
```

## 0. Basic installation
```{r, eval=FALSE}
# Development version from GitHub
devtools::install_github('filipezabala/voice')
# Stable from CRAN
install.packages('voice')
```
More details at https://github.com/filipezabala/voice.


## 1. Extract features
### 1.1 Load packages and audio files
```{r, message=FALSE, warning=FALSE}
# packs
library(voice)
library(tidyverse)

# get path to audio file
wavDir <- list.files(system.file('extdata', package = 'wrassp'),
                     pattern = glob2rx('*.wav'), full.names = TRUE)
```

### 1.2 Extract features
```{r, message=FALSE, warning=FALSE}
# minimal usage
M <- voice::extract_features(wavDir)
glimpse(M)
```


## 2. Tag
```{r, message=FALSE, warning=FALSE}
# creating Extended synthetic data
E <- dplyr::tibble(subject_id = c(1,1,1,2,2,2,3,3,3), wav_path = wavDir)
E

# minimal usage
voice::tag(E)

# canonical data
voice::tag(E, groupBy = 'subject_id')
```


## 3. Visualization
### 3.1 Get audio
```{r, message=FALSE, warning=FALSE}
url0 <- 'https://github.com/filipezabala/voiceAudios/raw/refs/heads/main/wav/doremi.wav'
download.file(url0, paste0(tempdir(), '/doremi.wav'), mode = 'wb')
voice::embed_audio(url0) # See https://github.com/mccarthy-m-g/embedr for more details.
```

### 3.2 Media data
```{r, message=FALSE, warning=FALSE}
M <- voice::extract_features(tempdir())
summary(M)
```

### 3.3 Plot
```{r, message=FALSE, warning=FALSE, fig.width=7.5, fig.height=4}
voice::piano_plot(M, 0) # f0
voice::piano_plot(M, 0:1) # f0 + f1
```

### 3.4 Assign notes
```{r, message=FALSE, warning=FALSE}
(f0_spn <- voice::assign_notes(M, fmt = 0, min_points = 22, min_percentile = .85)) # f0
(f1_spn <- voice::assign_notes(M, fmt = 1, min_points = 22, min_percentile = .85)) # f1
```

### 3.5 Sheet music
Must have [MuseScore](https://musescore.org/en) and [gm](https://flujoo.github.io/gm/articles/gm.html).

#### 3.5.1 Notes sequence of f0
```{r, message=FALSE, warning=FALSE}
library(gm)
line_0 <- gm::Line(as.character(f0_spn))
m0 <- gm::Music() +
  gm::Meter(4, 4) +
  line_0
gm::show(m0, to = c('score', 'audio'))
```


#### 3.5.2 Notes sequences of f0 and f1
```{r, message=FALSE, warning=FALSE}
line_0 <- gm::Line(as.character(f0_spn))
line_1 <- gm::Line(as.character(f1_spn))
m1 <- gm::Music() +
  gm::Meter(4, 4) +
  line_0 + line_1
gm::show(m1, to = c('score', 'audio'))
```

<!-- #### 3.6.3 Chords sequence using f0, f1 and f2  -->
<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- line_0 <- gm::Line(as.character(f0_spn)) -->
<!-- line_1 <- gm::Line(as.character(f1_spn)) -->
<!-- line_2 <- gm::Line(as.character(f2_spn)) -->
<!-- m2 <- gm::Music() +  -->
<!--   gm::Meter(4, 4) +  -->
<!--   line_0 + line_1 + line_2 -->
<!-- gm::show(m2, to = c('score', 'audio')) -->
<!-- ``` -->


## 4. Advanced installation
Python-based functions `diarize` and `extract_features` (when the latter is inferring `f0_praat` and `fmt_praat` features) require a configured Python environment.

### 4.1 Ubuntu
The following steps are used to fully configure `voice` on [Ubuntu 24.04 LTS (Noble Numbat)](https://ubuntu.com/blog/tag/ubuntu-24-04-lts). Reports of inconsistencies are welcome.

<!-- > Without the following Python items 3 and 10, you may run all the functions except `diarize` and `extract_features` calling `f0_praat` and `fmt_praat` features. These functions run respectively [pyannote-audio](https://github.com/pyannote/pyannote-audio) and [Parselmouth](https://github.com/YannickJadoul/Parselmouth). -->

#### 4.1.1. [Curl](https://curl.se/)
Command line tool and library for transferring data with URLs.
```bash
# installing dependencies
sudo apt-get update
sudo apt-get install -y libssl-dev autoconf libtool make
# installing curl
sudo apt install curl
# verify installation
curl --version
```

<!-- #### 4.1.2. [Git](https://git-scm.com/) -->
<!-- Git is a free and open source distributed version control system. -->
<!-- ```bash -->
<!-- sudo apt-get update -->
<!-- sudo apt-get install git-all -->
<!-- ``` -->

<!-- #### 4.1.3. [pip](https://pypi.org/project/pip/) -->
<!-- pip is the package installer for Python.  -->
<!-- ```bash -->
<!-- sudo apt-get update -->
<!-- sudo apt-get install python3-pip -->
<!-- pip3 --version -->
<!-- ``` -->

#### 4.1.2. [ffmpeg](https://ffmpeg.org/)
ffmpeg is a cross-platform solution to record, convert and stream audio and video.
```bash
sudo apt-get update
sudo apt-get install ffmpeg
```

#### 4.1.3. Audio drivers and extra packages
```bash
sudo apt-get update
sudo apt-get install portaudio19-dev libasound2-dev libfontconfig1-dev libmagick++-dev libxml2-dev libharfbuzz-dev libfribidi-dev libgdal-dev cmake cmake-doc ninja-build
```

#### 4.1.4. [MuseScore](https://musescore.org/)
MuseScore is an open source notation software.

```bash
sudo add-apt-repository ppa:mscore-ubuntu/mscore-stable
sudo apt-get update
sudo apt-get install musescore
```

#### 4.1.5. [R](https://www.r-project.org)
R is a free software environment for statistical computing and graphics. To find out your Ubuntu distribution use `lsb_release -a` at terminal.    
```bash
sudo sh -c 'echo "deb https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/" >> /etc/apt/sources.list.d/cran.list'
sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys E084DAB9 
sudo apt-key adv --keyserver keyserver.ubuntu.com --recv-keys 51716619E084DAB9
gpg -a --export E084DAB9 | sudo apt-key add -

sudo add-apt-repository ppa:c2d4u.team/c2d4u4.0+

sudo apt-get update && sudo apt-get upgrade
sudo apt-get install r-base r-base-dev
```

#### 4.1.6. [RStudio](https://posit.co/download/rstudio-desktop/)
RStudio is an Integrated Development Environment (IDE) for R. Check for updates [here](https://posit.co/download/rstudio-desktop/).
```bash
sudo apt-get update
sudo apt-get install gdebi-core
wget https://download1.rstudio.org/electron/jammy/amd64/rstudio-2025.05.0-496-amd64.deb
sudo gdebi rstudio-2025.05.0-496-amd64.deb
```

#### 4.1.9. R packages
"Packages are the fundamental units of reproducible R code." [Hadley Wickham and Jennifer Bryan](https://r-pkgs.org/). The installation may take several minutes. At terminal run:
```bash
sudo R
```

Running R as super user paste the following, row by row:
```r
packs <- c('audio','reticulate','R.utils','seewave','tidyverse','tuneR','wrassp')
install.packages(packs, dep = TRUE)
update.packages(ask = FALSE)
devtools::install_github('egenn/music')
devtools::install_github('flujoo/gm')
```
To configure the `gm` package.
```r
usethis::edit_r_environ()
```

Add the line `MUSESCORE_PATH=/usr/bin/mscore` to `/root/.Renviron` file. To exit use `:wq` at VI. Save and restart the R/RStudio session.


#### 4.1.10. [Miniconda](https://docs.conda.io/en/latest/miniconda.html)
Miniconda is a free minimal installer for [conda](https://docs.conda.io/), an open source package, dependency and environment management system for any language—Python, R, Ruby, Lua, Scala, Java, JavaScript, C/ C++, FORTRAN and more, that runs on Windows, macOS and Linux.   
Follow the instructions at https://docs.conda.io/en/latest/miniconda.html. 

At terminal:
```bash
cd ~/Downloads/
wget -r -np -k https://repo.anaconda.com/miniconda/Miniconda3-latest-Linux-x86_64.sh
cd repo.anaconda.com/miniconda/
bash Miniconda3-latest-Linux-x86_64.sh
```
Do you accept the license terms? [yes|no] `yes`.

Miniconda3 will now be installed into this location: /home/user/miniconda3 [ENTER]

You can undo this by running `conda init --reverse $SHELL`? `yes`

Do you wish the installer to initialize Miniconda3 by running conda init? `yes`.

Close and reopen terminal.

```bash
conda update -n base -c defaults conda
```

The following packages will be INSTALLED/REMOVED/UPDATED/DOWNGRADED:... Proceed ([y]/n)? `y`

```bash
conda create -n pyvoice python=3.12
```

The following (NEW) packages will be downloaded/INSTALLED:... Proceed ([y]/n)? `y`

```bash
conda activate pyvoice
pip install -r https://raw.githubusercontent.com/filipezabala/voice/master/requirements.txt
```


### 4.2 MacOS
The following steps are used to fully configure `voice` on [MacOS Sonoma](https://www.apple.com/macos/macos-sequoia/) (Link to MacOS Sequoia). Reports of inconsistencies are welcome.

<!-- Hardware   -->
<!--  . MacBook Air (13-inch, 2020)   -->
<!--  . Chip Apple M1 -->
<!--  . Memory 8GB LPDDR4 -->

#### 4.2.1. [Homebrew](https://brew.sh/)
Install Homebrew, 'The Missing Package Manager for macOS (or Linux)' and remember to `brew doctor` eventually. At terminal (`command + space 'terminal'`) run:
```bash
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
```

#### 4.2.2. [wget](https://www.gnu.org/software/wget/)
GNU Wget is a free software package for retrieving files using HTTP, HTTPS, FTP and FTPS, the most widely used Internet protocols. It is a non-interactive commandline tool, so it may easily be called from scripts, cron jobs, terminals without X-Windows support, etc.
```bash
brew install wget
```

#### 4.2.3. [Python](https://www.python.org/)
Python is a programming language that integrate systems. According to [this](https://github.com/Homebrew/homebrew-core/issues/62911) post, it is recommended to install Python 3.8 and 3.9 and make it consistent.
```bash
brew install python@3.12
python3 --version 
pip3 --version
```

#### 4.2.4. [ffmpeg](https://ffmpeg.org/)
ffmpeg is a cross-platform solution to record, convert and stream audio and video. The installation may take several minutes.
```bash
brew install ffmpeg
```

#### 4.2.5. [XQuartz](https://www.xquartz.org/)
The XQuartz project is an open-source effort to develop a version of the [X.Org X Window System](https://www.x.org/wiki/) that runs on macOS.

 - Download and run https://github.com/XQuartz/XQuartz/releases/download/XQuartz-2.8.5/XQuartz-2.8.5.pkg
 - Will take around 320 MB of disk space   
 - Send XQuartz-2.8.5.dmg to Trash    


#### 4.2.6. [MacPorts](https://guide.macports.org/chunked/installing.macports.html)
Follow the instructions from https://guide.macports.org/chunked/installing.macports.html.

<!-- ```bash -->
<!-- curl -O https://distfiles.macports.org/MacPorts/MacPorts-2.10.7.tar.bz2 -->
<!-- tar xf MacPorts-2.10.7.tar.bz2 -->
<!-- cd MacPorts-2.10.7 -->
<!-- ./configure -->
<!-- make -->
<!-- sudo make install -->
<!-- cd ../; rm -rf MacPorts-2.10.7* -->
<!-- sudo nano /etc/paths -->
<!-- ``` -->
<!-- Add the following lines at the end of the file. -->
<!-- ```bash -->
<!-- /opt/local/bin -->
<!-- /opt/local/sbin -->
<!-- ``` -->
<!-- Use Ctrl+O and return to save the file and Ctrl+X to close nano editor. Reboot your terminal and use -->
<!-- ```bash -->
<!-- sudo port -v selfupdate -->
<!-- sudo port upgrade outdated -->
<!-- ``` -->

#### 4.2.7. [tcllib](https://ports.macports.org/port/tcllib/)

```bash
sudo port selfupdate && sudo port upgrade tcllib
sudo port install tcllib
```

#### 4.2.8. [MuseScore](https://musescore.org/)
MuseScore is an open source notation software.

 - Download and run https://musescore.org/en/download/musescore.dmg  
 - Drag MuseScore 4 to Applications folder   
 - Will take around 320 MB of disk space   
 - Unmount MuseScore-4.2.1 virtual disk and send MuseScore-4.2.1.240230937.dmg to Trash   


#### 4.2.9. [R](https://www.r-project.org)
R is a free software environment for statistical computing and graphics.  

 - Download and run the pkg file according to you architecture from https://cloud.r-project.org/bin/macosx/
 - Will take around 180 MB of disk space   


#### 4.2.10. [RStudio](https://posit.co/download/rstudio-desktop/)
RStudio is an Integrated Development Environment (IDE) for R.

 - Download and run https://download1.rstudio.org/electron/macos/RStudio-2025.05.0-496.dmg
 - Drag RStudio to Applications folder   
 - Will take around 770 MB of disk space   
 - Unmount RStudio virtual disk and send RStudio-2025.05.0-496.dmg to Trash   
 - Type `command + space 'rstudio'`   
 - Tools > Global Options... > Appearance > Merbivore (Restart required)   


#### 4.2.11. R packages
"Packages are the fundamental units of reproducible R code." [Hadley Wickham and Jennifer Bryan](https://r-pkgs.org/). Type `command + space 'terminal'`   
```bash
sudo R
```

Running R as super user paste the following, one line at a time.
```r
packs <- c('audio','reticulate','R.utils','seewave','tidyverse','tuneR','wrassp')
install.packages(packs, dep = TRUE)
update.packages(ask = FALSE)
devtools::install_github('egenn/music')
devtools::install_github('flujoo/gm')
```

#### 4.2.12. [Miniconda](https://docs.conda.io/en/latest/miniconda.html)
Miniconda is a free minimal installer for [conda](https://docs.conda.io/), an open source package, dependency and environment management system for any language—Python, R, Ruby, Lua, Scala, Java, JavaScript, C/ C++, FORTRAN and more, that runs on Windows, macOS and Linux. 

For 64-bit version use

```bash
cd ~/Downloads
wget -r -np -k https://repo.anaconda.com/miniconda/Miniconda3-latest-MacOSX-x86_64.sh
cd repo.anaconda.com/miniconda/
bash Miniconda3-latest-MacOSX-x86_64.sh
```

For M1 version use

```bash
cd ~/Downloads
wget -r -np -k https://repo.anaconda.com/miniconda/Miniconda3-latest-MacOSX-arm64.sh
cd repo.anaconda.com/miniconda/
bash Miniconda3-latest-MacOSX-arm64.sh
```

In order to continue the installation process, please review the license
agreement. Please, press ENTER to continue `ENTER`.

You can undo this by running `conda init --reverse $SHELL`? `yes`

Close and reopen terminal.

```bash
export PATH="~/miniconda3/bin:$PATH"
conda update -n base -c defaults conda
```

The following packages will be INSTALLED/REMOVED/UPDATED/DOWNGRADED:... Proceed ([y]/n)? `y`

```bash
conda create -n pyvoice python=3.12
```
The following (NEW) packages will be downloaded/INSTALLED:... Proceed ([y]/n)? `y`

Close and reopen terminal.

```bash
conda activate base
conda activate pyvoice
pip install -r https://raw.githubusercontent.com/filipezabala/voice/master/requirements.txt
```

## 5. Diarize

```{r}
# download
url0 <- 'https://github.com/filipezabala/voiceAudios/raw/main/wav/sherlock0.wav'
wavDir <- normalizePath(tempdir())
download.file(url0, paste0(wavDir, '/sherlock0.wav'), mode = 'wb')
```
Diarization can be performed to detect speaker segments (i.e., 'who spoke when').
```{r}
# diarize
voice::diarize(fromWav = wavDir, toRttm = wavDir, token = 'YOUR_TOKEN')
```

The `voice::diarize()` function creates Rich Transcription Time Marked (RTTM)[^rttm] files, space-delimited text files containing one turn per line defined by NIST - National Institute of Standards and Technology. The RTTM files can be read using `voice::read_rttm()`.

[^rttm]: See Appendix	C at https://www.nist.gov/system/files/documents/itl/iad/mig/KWS15-evalplan-v05.pdf.

```{r}
# read_rttm
(rttm <- voice::read_rttm(wavDir))
```

Finally, the audio waves can be automatically segmented.
```{r}
# split audio wave
voice::splitw(fromWav = wavDir, fromRttm = wavDir, to = wavDir)
dir(wavDir, pattern = '.[Ww][Aa][Vv]$')
```
